---
title: "Classification with AddiVortes"
author: "John Paul Gosling and Adam Stone"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Classification with AddiVortes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4.5,
  fig.align = "center"
)
```

This vignette shows how `AddiVortes()` fits **binary** and **multi-category** classification models. The task is chosen automatically from the response `y`. Binary classification uses a probit link with Albert-Chib latent variables. Multi-category classification uses the same idea with \(K-1\) independent latent ensembles (the first class is the reference level).

```{r, message=FALSE, warning=FALSE}
library(AddiVortes)
```

## How the response is interpreted

`AddiVortes()` looks only at `y`:

- A **factor**, **character** or **logical** vector is classification. Two levels give a binary model; three or more give a multinomial model.
- A **numeric** vector with exactly two unique values in `{0, 1}` is binary classification (1 is the event class).
- Any other numeric vector is Gaussian regression, as in previous versions of the package.

The first factor level is the reference class, matching `glm()`. For numeric 0/1 data, 0 is the reference. Character vectors are converted with `factor()`, so levels are alphabetical.

Missing values in `y` are not allowed, and a classification response must have at least two classes.

## Binary classification

We simulate a two-dimensional problem whose labels follow a simple linear boundary.

```{r}
set.seed(2026)
n <- 120
x <- data.frame(
  x1 = runif(n, -1, 1),
  x2 = runif(n, -1, 1)
)
y <- factor(ifelse(x$x1 + x$x2 > 0, "active", "inactive"),
            levels = c("inactive", "active"))

fit_bin <- AddiVortes(
  y, x,
  m = 200,
  totalMCMCIter = 2000,
  mcmcBurnIn = 200,
  showProgress = FALSE
)

fit_bin$task
fit_bin$classLevels
fit_bin$inSampleAccuracy
```

`predict(..., type = "response")` returns the posterior mean of \(\Phi(G(x))\), the probability of the event class (`active` here). `type = "class"` returns labels, and `type = "quantile"` gives credible intervals for those probabilities.

```{r}
p_hat <- predict(fit_bin, x, type = "response", showProgress = FALSE)
y_hat <- predict(fit_bin, x, type = "class", showProgress = FALSE)
p_int <- predict(fit_bin, x,
  type = "quantile",
  quantiles = c(0.05, 0.95),
  showProgress = FALSE
)

mean(y_hat == y)
range(p_hat)
head(cbind(prob = round(p_hat, 3), lower = round(p_int[, 1], 3),
           upper = round(p_int[, 2], 3), class = as.character(y_hat)))
```

Observations with probability near 0.5 are the most uncertain. The 90% intervals typically cover 0.5 for those points.

```{r}
plot(p_hat, as.integer(y) - 1,
     xlab = "Predicted P(active)",
     ylab = "Observed class",
     pch = 19, col = "steelblue")
abline(v = 0.5, lty = 2, col = "grey40")
```

## Multi-category classification

With three or more classes, AddiVortes fits \(K-1\) latent ensembles. The argument `m` is the number of tessellations **per latent**, so a three-class model with `m = 15` uses 30 tessellations in total.

```{r}
set.seed(2026)
n3 <- 90
x3 <- data.frame(
  x1 = rnorm(n3),
  x2 = rnorm(n3)
)
y3 <- cut(x3$x1, breaks = c(-Inf, -0.5, 0.5, Inf), labels = c("low", "mid", "high"))

fit_multi <- AddiVortes(
  y3, x3,
  m = 200,
  totalMCMCIter = 2000,
  mcmcBurnIn = 200,
  showProgress = FALSE
)

fit_multi$task
fit_multi$nLatents
fit_multi$inSampleAccuracy
```

`type = "response"` is now an \(n \times K\) matrix of class probabilities. Rows sum to 1. `type = "class"` returns the class with the largest probability.

```{r}
p3 <- predict(fit_multi, x3, type = "response", showProgress = FALSE)
cls3 <- predict(fit_multi, x3, type = "class", showProgress = FALSE)

head(round(p3, 3))
mean(abs(rowSums(p3) - 1) < 1e-8)
table(predicted = cls3, observed = y3)
```

## Notes on the classification prior

On the latent probit scale the residual variance is fixed at \(\sigma = 1\), so `nu`, `q` and `InitialSigma` are ignored. Cell means use

\[
\sigma_\mu = \frac{3}{k\sqrt{m}},
\]

which keeps \(G(x)\) typically inside \((-3, 3)\) and shrinks class probabilities towards 0.5 (or towards equal class probabilities in the multinomial case). The default `k = 3` is a reasonable starting point.

The response itself is not scaled. Covariates are still centred and scaled exactly as in regression.

For further details of the binary model see Stone, Ogundimu and Gosling (2026). The multi-category construction follows the latent-utility setup of Kindo, Wang and Peña (2016), with independent latents (\(\Sigma = I\)).
